---
title: "Notes for Group"
author: "Lorenzo Booth (asmtry)"
output: html_document
---
What follows is just a running log for early 'development'.

# Merced Watershed Stations
# Crane Flats
  * Water Years: 2001-2013
  * Properties: date_time, speed, direc, temp humidity, rad
## Jerseydale
  * Water Years: 2002-2013 (!!! Starts at 2002!)
  * Properties: date_time, speed, direc, temp humidity, rad
## Metcalf Gap
  * Water Years: 2001-2013
  * Properties: date_time, speed, direc, temp humidity, rad

# Tuolumne Watershed Stations
## Gaylord Meadows
  * Water Years: 2001-2011 (!!! Ends at 2011!)
  * Properties: date_time, speed, direc, temp humidity, rad
## Mt. Elizabeth
  * Water Years: 2001-2013
  * Properties: date_time, speed, direc, temp humidity, rad
## White Wolf
  * Water Years: 2001-2013
  * Properties: date_time, speed, direc, temp humidity, rad

# Treatments applied so far, using Crane_Flat_RAWS as an example
The part of the function that compiles specially named .csv's and puts them in one data frame.
```{r}
#Part of readRAWS, use that instead
#Or run in folder containing .csvs
#All csvs must have the same number of columns and the same column names
csvfiles = list.files()
YOW_precip_inc_01_13<- do.call(rbind, lapply(csvfiles, function(x) {read.csv(x, header=TRUE)}))
```

It seems that -999 values are the straight-up missing values.
```{r}
MCD_Crane_Flats_01_13[MCD_Crane_Flats_01_13 == -999] = NA
```


Re-class the `date_time` column into one that can be parsed with the POSIXct family of functions.
```{r}
#Given an example date format 2001-10-01 10:00:00
MCD_Crane_Flats_01_13$date_time = as.POSIXct(MCD_Crane_Flats_01_13$date_time,format="%Y-%m-%d %H:%M:%S")
```

Create sample categories for the purpose of doing monthly averages for example.
```{r}
MCD_Crane_Flats_01_13$monthYr = strftime(result$date_time, format="%Y/%m")
```

**NOTE THAT NA VALUES ARE NOT REMOVED YET. This may cause certain functions to fail in mysterious ways**
# Treatments not applied yet that may be of interest, using Crane_Flat_RAWS as an example

Lets plot temperature to look for potentially funky values and gaps in our information.
```{r}
plot(MCD_Crane_Flats_01_13$Temp, pch=16, cex=0.4)
# Oh dear.
```


Let's take the averages of each property for each month and try to graph it.
```{r}
# Note, aggregate omits NAs by default. Not every function plays this nice.
CF_Monthly_Means = aggregate(cbind(Speed,Direc,Temp,Humidty,Rad.) ~ monthYr, data=MCD_Crane_Flats_01_13, mean)
plot(CF_Monthly_Means$Temp, pch=16, cex=0.4)
lines(CF_Monthly_Means$Temp)
# Yay! We see something that is cyclic as expected!
```


